\sysfull

### 2018-04-15 paper/intro.tex

Advances in processor efficiency along with the development of
energy-harvesting systems has created a new category of devices that require
neither a battery nor a tethered power
supply~\citep{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These
devices operate using ambient energy, such as radio frequency
transmissions~\citep{rf_powered_computing_gollakota_2014},
light~\citep{margolies_infocom_2016,margolies_tosn_2016}, and
vibration~\citep{gorlatova_sigmetrics_2014}. Incorporating compute, storage,
sensing, and communication hardware~\citep{wisp5,moo,capybara}, such devices are a
promising technology for use in the Internet of Things~\citep{ku_cst_2016},
in-body~\citep{nadeau_naturebio_2017} and
on-body~\citep{bandodkar_electroanalysis_2015} medical systems, and
energy-harvesting nano-satellites~\citep{kicksat,capybara}.

Energy-harvesting devices create unique challenges because they operate {\em
intermittently} when energy is
available~\citep{hicks_isca_2017,lucia_snapl_2017}. An energy-harvesting device
buffers energy in a small storage capacitor~\citep{gorlatova_tmc_2013,gunduz_commag_2014} and operates when a
threshold amount of energy has accumulated. Harvestable energy sources are low-power (e.g., nW to $\mu$W) compared to a platform's operating
power level (hundreds of $\mu$W to mW). A device operates briefly until it depletes its buffered energy, after which, the device shuts
down and recharges to operate again later---corresponds to {\em intermittent execution model}~\citep{dino,lucia_snapl_2017} composed of
operation-power failure-restart cycles. As an example, recharge times may be
tens of seconds in radio frequency powered medical device~\cite[Fig.
3c]{nadeau_naturebio_2017}.  The recharge and discharge time---which corresponds to the device's inactive and active time---varies with the energy buffering capacitor's~\cite{capybara} and some devices fail and restart operating $\approx$10 to
$\approx$100 times per second~\citep{tan_infocom_2016,mementos,nvp}.

@@

\textbf{Data Consistency in Intermittent Computing.}  
%Software in an energy-harvesting system operates according to the {\em intermittent execution model}~\citep{dino,lucia_snapl_2017}, which corresponds to the operation-failure-restart cycles of an energy-harvesting device. In an intermittent software execution, an {\em operating period} proceeds for anarbitrary duration (dictated by energy availability) before being interruptedby a {\em failure period}. In a failure period, 
Upon power failures, a device loses the volatile
state in its registers, stack, SRAM, and retains the state of any non-volatile
memory, such as FRAM or Flash. While capturing periodic
checkpoints~\citep{mementos,quickrecall} and sleep
scheduling~\citep{dewdrop,hibernus,hibernusplusplus} help preserve execution
progress, failures can leave non-volatile state incorrectly, partially updated,
and may leave checkpointed volatile state inconsistent with non-volatile state.
These inconsistencies cause intermittent execution to deviate from
continuously-powered behavior, often leading to an unrecoverable
application failure~\citep{dino,edb}. Prior work developed two main approaches to dealing with data inconsistency for
intermittently-powered devices: (i) \emph{software-based programming and
execution models}~\citep{dino,ratchet,chain,alpaca} and (ii)
\emph{hardware-based computer architecture
support}~\citep{hicks_isca_2017,idetic,nvp}.  Complex architectural changes are
expensive to design, verify, and manufacture.  New architectures are also
inapplicable to existing systems~\citep{hicks_isca_2017,nvp}. Software
approaches are simpler and applicable to existing devices today, but prior
software approaches have a number of limitations that are the focus of the work
in this paper.  A key limitation of prior software approaches that we address in this paper is the {\em inflexibility} of a program statically decomposed into tasks.

@@

A savvy programmer may thus create very large tasks in an attempt to minimize
overhead by minimizing the number of task transitions.  However, a very large
task may require more energy to complete than a device's fixed hardware energy
buffer can hold.  Such a large task fails to completely execute assuming only
buffered energy is available (i.e., assuming the worst case).  While some
environments may have sufficient harvestable energy to replenish a device's
capacitor during operation, prolonging the execution period and enabling the
task to complete, assuming cooperation from the physical environment to avoid
non-termination is a risky programming proposition.  To eliminate this risk,
existing systems require the programmer or compiler to decompose a  program
into small tasks, all of which complete using only buffered energy.  These
constraints on task sizing leave an unsatisfying dilemma: large, efficient
tasks that risk non-termination, or small tasks that are guaranteed to
complete, but incur a high task transition and commit overhead.  In this work,
we propose a third way, using a novel technique called {\em adaptive execution by dynamic task
coalescing/splitting} that efficiently executes small tasks, avoiding unnecessary overheads while avoiding the risk of non-termination.

@@

This paper introduces a new system called \sys, which executes a task-based
intermittent program by dynamically coalescing/splitting its tasks.
\sys accepts any static decomposition (i.e., from the programmer or from a compiler) 
and coalesces/splits its tasks at runtime. Two consecutive, coalesced tasks execute  
with no commit or task transition overhead between them, instead performing
task-end commit actions at the end second task only.
%
When there is sufficient energy to execute both tasks, two distinct, small
tasks are effectively executed as a single large task, preserving the atomicity
of both.  If power fails during a coalesced task, execution restarts from the
{\em first} of the coalesced tasks.  Figure~\ref{fig:coalesce} (top)
illustrates how dynamic coalescing changes the earlier execution and highlights
the change in memory model necessary to not break atomicity. On the other hand, a large task that requires more energy to complete than a device's fixed energy buffer is split by \sys to enable task termination, as depicted in Figure~\ref{fig:coalesce} (bottom).

@@

%variables in non-volatile memory at \emph{run-time} that can
%break the atomicity of the \emph{virtual task}. Consider the
%example depicted in Fig.~\ref{fig:virtualization}: Tasks 1,
%2 and 3 being executed consecutively. All tasks in this
%example are atomic since they do not have WAR dependency on
%the persistent variables they are accessing, e.g. \emph{x}
%is only read and \emph{y} is only written within Task 1.
%Now, suppose three tasks have been virtualized into a single
%one at run-time, namely Task 4: since \emph{x} is now first
%\emph{read} and then \emph{written}, a WAR dependency on
%\emph{x} is introduced dynamically at
%run-time---unfortunately Task 4 is \emph{no more atomic}
%since its re-execution will not always produce the same
%results.
\textbf{Challenges.} While a compiler can use static data privatization and commit instrumentation
(i.e., redo-logging) to eliminate statically identifiable, inter-task data
dependences~\citep{alpaca}, a \emph{dynamic} dependence between two coalesced
tasks requires dynamic privatization and commit actions. 
%
To ensure consistency and respect inter-task dependences, \sys uses a novel
approach to \emph{memory virtualization} that buffers non-volatile variable
updates in volatile memory during coalesced task execution, before committing
them to non-volatile memory at the dynamic task boundary.
%
%Therefore, we require a \emph{new execution
%model} that keeps each virtual task atomic by committing the
%modified persistent variables at the boundary defined
%dynamically at run-time---keeping the non-volatile memory
%unmodified upon a power-interrupt and preserving its
%consistency.
A static task decomposition model assumes that each
single task can execute to completion.  If the hardware energy buffer provides
inadequate energy to execute each single task to completion, a program will not
terminate~\cite{cleancut_2018}. To avoid non-termination under adversarial 
energy conditions, \sys uses a timer-based {\em partial task commit} mechanism.
Partial commit avoids non-termination by committing the intermediate state of a long-running
task that has repeatedly failed and restarted.  Partial commit violates
task atomicity, but preserves forward progress; if a programmer knows that
task atomicity is crucial to correctness, they can disable partial commit
instead risking non-termination.

@@

\begin{itemize}
\item The first {\em adaptive} task-based intermittent execution model that dynamically coalesces tasks to avoid unnecessary commit overhead. 
\item The first software memory virtualization mechanism for an intermittent system that \sys uses to preserve the atomicity of coalesced tasks.
\item A fall-back dynamic task splitting mechanism that preserves forward progress despite executing tasks that are too large for a device's energy supply.
\item A fully-realized prototype of \sys that runs on real energy-harvesting devices, to be released as open source after publication\footnote{The \sys repository will be accessible via~\cite{coala_website} immediately after publication.}. 
\item An evaluation directly comparing \sys to a state of the art task-based intermittent programming and execution model from prior work, showing that on a suite of benchmarks from the literature and several new workloads, \sys often has higher performance and is more flexible to varied energy conditions. 
\end{itemize}

>>>>

Advances in processor efficiency along with the development of
energy-harvesting systems have created a new category of devices that require
neither a battery nor a tethered power
supply~\citep{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These
devices operate using ambient energy, such as radio frequency
transmissions~\citep{rf_powered_computing_gollakota_2014},
light~\citep{margolies_infocom_2016,margolies_tosn_2016}, and
vibration~\citep{gorlatova_sigmetrics_2014}. Incorporating compute, storage,
sensing, and communication hardware~\citep{wisp5,moo,capybara}, such devices are a
promising technology for use in the Internet of Things~\citep{ku_cst_2016},
in-body~\citep{nadeau_naturebio_2017} and
on-body~\citep{bandodkar_electroanalysis_2015} medical systems, and
energy-harvesting nano-satellites~\citep{kicksat,capybara}.

Energy-harvesting devices create unique challenges because they operate {\em
intermittently} when energy is
available~\citep{hicks_isca_2017,lucia_snapl_2017}. An energy-harvesting device
buffers energy in a small storage capacitor~\citep{gorlatova_tmc_2013,gunduz_commag_2014} and operates when a
threshold amount of energy has accumulated. Harvestable energy sources are low-power (e.g., nW to $\mu$W) compared to a platform's operating
power level (hundreds of $\mu$W to mW). A device operates briefly until it depletes its buffered energy, after which, the device shuts
down and recharges to operate again later---corresponds to {\em intermittent execution model}~\citep{dino,lucia_snapl_2017} composed of
operation-power failure-restart cycles. As an example, recharge times may be
tens of seconds in radio frequency powered medical device~\cite[Fig.
3c]{nadeau_naturebio_2017}. The recharge and discharge time---which corresponds to the device's inactive and active time---varies with the energy buffering capacitor~\cite{capybara} and some devices fail and restart operating $\approx$10 to
$\approx$100 times per second~\citep{tan_infocom_2016,mementos,nvp}.

@@

\textbf{Data Consistency in Intermittent Computing.}  
%Software in an energy-harvesting system operates according to the {\em intermittent execution model}~\citep{dino,lucia_snapl_2017}, which corresponds to the operation-failure-restart cycles of an energy-harvesting device. In an intermittent software execution, an {\em operating period} proceeds for anarbitrary duration (dictated by energy availability) before being interruptedby a {\em failure period}. In a failure period, 
Upon power failures, a device loses the volatile
state in its registers, stack, SRAM, and retains the state of any non-volatile
memory, such as FRAM or Flash. While capturing periodic
checkpoints~\citep{mementos,quickrecall} and sleep
scheduling~\citep{dewdrop,hibernus,hibernusplusplus} help preserve execution
progress, failures can leave non-volatile state incorrectly, partially updated,
and may leave checkpointed volatile state inconsistent with non-volatile state.
These inconsistencies cause intermittent execution to deviate from
continuously-powered behavior, often leading to an unrecoverable
application failure~\citep{dino,edb}. Prior work developed two main approaches to deal with data inconsistency for
intermittently-powered devices: (i) \emph{software-based programming and
execution models}~\citep{dino,ratchet,chain,alpaca} and (ii)
\emph{hardware-based computer architecture
support}~\citep{hicks_isca_2017,idetic,nvp}. Complex architectural changes are
expensive to design, verify, and manufacture. New architectures are also
inapplicable to existing systems~\citep{hicks_isca_2017,nvp}. Software
approaches are simpler and applicable to existing devices today, but prior
software approaches have a number of limitations that are the focus of the work
in this paper. A key limitation of prior software approaches that we address in this paper is the {\em inflexibility} of a program statically decomposed into tasks.

@@

A savvy programmer may thus create very large tasks in an attempt to reduce
overhead by minimizing the number of task transitions. However, a very large
task may require more energy to complete than a device's fixed hardware energy
buffer can hold. Such a large task fails to completely execute assuming only
buffered energy is available (i.e., assuming the worst case). While some
environments may have sufficient harvestable energy to replenish a device's
capacitor during operation, prolonging the execution period and enabling the
task to complete, assuming cooperation from the physical environment to avoid
non-termination is a risky programming proposition. To eliminate this risk,
existing systems require the programmer or compiler to decompose a program
into small tasks, all of which complete using only buffered energy. These
constraints on task sizing leave an unsatisfying dilemma: large, efficient
tasks that risk non-termination, or small tasks that are guaranteed to
complete, but incur a high task transition and commit overhead. In this work,
we propose a third way, using a novel technique called {\em adaptive execution by dynamic task
coalescing/splitting} that efficiently executes small tasks, trimming unnecessary overheads while avoiding the risk of non-termination.

@@

This paper introduces a new system called \sys, which executes a task-based
intermittent program by dynamically coalescing/splitting its tasks.
\sys accepts any static decomposition (i.e., from the programmer or from a compiler) 
and coalesces/splits its tasks at runtime. Two consecutive, coalesced tasks execute  
with no commit or task transition overhead between them, instead performing
task-end commit actions at the end of the second task only.
%
When there is sufficient energy to execute both tasks, two distinct, small
tasks are effectively executed as a single large task, preserving the atomicity
of both. If power fails during a coalesced task, execution restarts from the
{\em first} of the coalesced tasks. Figure~\ref{fig:coalesce} (top)
illustrates how dynamic coalescing changes the earlier execution and highlights
the change in memory model necessary to not break atomicity. On the other hand, a large task that requires more energy to complete than a device's fixed energy buffer is split by \sys to enable task termination, as depicted in Figure~\ref{fig:coalesce} (bottom).

@@

%variables in non-volatile memory at \emph{run-time} that can
%break the atomicity of the \emph{virtual task}. Consider the
%example depicted in Fig.~\ref{fig:virtualization}: Tasks 1,
%2 and 3 being executed consecutively. All tasks in this
%example are atomic since they do not have WAR dependency on
%the persistent variables they are accessing, e.g. \emph{x}
%is only read and \emph{y} is only written within Task 1.
%Now, suppose three tasks have been virtualized into a single
%one at run-time, namely Task 4: since \emph{x} is now first
%\emph{read} and then \emph{written}, a WAR dependency on
%\emph{x} is introduced dynamically at
%run-time---unfortunately Task 4 is \emph{no more atomic}
%since its re-execution will not always produce the same
%results.
\textbf{Challenges.} While a compiler can use static data privatization and commit instrumentation
(i.e., redo-logging) to eliminate statically identifiable, inter-task data
dependences~\citep{alpaca}, a \emph{dynamic} dependence between two coalesced
tasks requires dynamic privatization and commit actions. 
%
To ensure consistency and respect inter-task dependences, \sys uses a novel
approach to \emph{memory virtualization} that buffers non-volatile variable
updates in volatile memory during a coalesced task execution, before committing
them to non-volatile memory at the dynamic task boundary.
%
%Therefore, we require a \emph{new execution
%model} that keeps each virtual task atomic by committing the
%modified persistent variables at the boundary defined
%dynamically at run-time---keeping the non-volatile memory
%unmodified upon a power-interrupt and preserving its
%consistency.
A static task decomposition model assumes that each
single task can execute to completion. If the hardware energy buffer provides
inadequate energy to execute each single task to completion, a program will not
terminate~\cite{cleancut_2018}. To avoid non-termination under adversarial
energy conditions, \sys uses a timer-based {\em partial task commit} mechanism.
Partial execution avoids non-termination by committing the intermediate state of a long-running
task that has repeatedly failed and restarted. Partial execution violates
task atomicity, but preserves forward progress; if a programmer knows that
task atomicity is crucial to correctness, they can disable partial execution around a piece of code.

@@

\begin{itemize}
\item The first {\em adaptive} task-based intermittent execution model that dynamically coalesces tasks to avoid unnecessary commit overhead. 
\item The first software memory virtualization mechanism for an intermittent system that \sys uses to preserve the atomicity of coalesced tasks.
\item A fall-back dynamic task splitting mechanism that preserves forward progress despite executing tasks that are too large for a device's energy supply.
\item A fully-realized prototype of \sys that runs on real energy-harvesting devices, to be released as open source after publication\footnote{The \sys repository will be accessible via~\cite{coala_website} immediately after publication.}. 
\item An evaluation directly comparing \sys to a state of the art task-based intermittent programming and execution model from prior work, showing that on a suite of benchmarks from the literature and several new workloads, \sys often has higher performance and is more flexible to diverse energy conditions. 
\end{itemize}

### 2018-04-15 paper/partialcommit.tex

\begin{figure}
	\caption{\textcolor{red}{Figure to describe partial commit is needed}}
\end{figure}

@@

A partial commit saves the stack, registers and the volatile pages to the non-volatile memory.

\sys starts with a big random guess and down-scales it using a binary search until it succeeds in preserving the partial execution progress. The obtained value is the initial seed for the next partial commit procedure.

@@

\subsection{Why partial commit is not \sys's default behavior}

\emph{Static deployment of partial commit stubs} This approach reduces programmer effort---they still need to apply \sys's API and distribute trigger points for partial commits---but it introduces an additional overhead by saving the stack and the register file each time it commits the pages.

\emph{Dynamic deployment of partial commit stubs:} This approach does not rely on any information from the programmer to trigger the commit process. However, it utilizes a trial and error approach to optimize the commit rate over the entire program execution time. Therefore, it suffers from a large cumulative re-execution penalty.

>>>>

We proceed with the second task adaptation feature of \sys, namely reduction of the task size boundary to enable execution of a program on any energy buffer size.

\begin{figure}
	\caption{\todo{Figure to describe partial commit is needed}{Sinan, Carlo}}
\end{figure}

@@

A partial commit saves the stack, registers and the volatile pages to the non-volatile memory. \sys starts with a big random guess and down-scales it using a binary search until it succeeds in preserving the partial execution progress. The obtained value is the initial seed for the next partial commit procedure.

@@

\todo{Proof of the maximum granularity of task downscaling - show it is limited by the clock}{Carlo}

\subsubsection{Why Task Downsizing is not \sys's Default Behavior}

Task downsizing is the core novelty of \sys when it comes to code portability and re-usability. However, it has limitations, which force  it to be a non-default \sys behavior.

\textbf{Static deployment of partial commit stubs.} This approach reduces programmer effort, although the programmer still needs to apply \sys's API and distribute trigger points for partial commits. On the other hand, \sys' downsizing introduces an additional overhead by saving the stack and the register file each time it commits the pages.

\textbf{Dynamic deployment of partial commit stubs.} \sys downsizing does not rely on any information from the programmer to trigger the commit process. However, \sys utilizes a trial and error approach to optimize the commit rate over the entire program execution time. Therefore, it suffers from a large cumulative re-execution penalty.
